Adaptive Distance Metric Selection for Supervised Classification

### 2016-10-19 latex/Adaptive_distance_metric_notes.tex
sync with emails

@@

Suppose we are given $N$ points $\v{v}_i \in \Re^p$ with class membership $\mc{C}_i$ for $i \in \mc{N} = \{1,\dots,N\}$.
Our goal is to find a distance metric $\D: \Re^p \mapsto \Re$ such that the following conditions hold:
\begin{enumerate}
\item $\D(\v{v}_i,\v{v}_j) = \v{a}^{\top}(\v{v}_i - \v{v}_j) + (\v{v}_i - \v{v}_j)^{\top} \v{B}(\v{v}_i - \v{v}_j) \quad \forall i,j \in \mc{N}$, for some $\v{a} \in \Re^p$ and $\v{B} \in \Re^{p \times p}$ and symmetric
\item $\D(\v{v}_i,\v{v}_j) \geq 0 \quad \forall i,j \in \mc{N}$
\item $\min_{j \in \mc{C}_i} \D(\v{v}_i,\v{v}_j) < \min_{k \notin \mc{C}_i} \D(\v{v}_i,\v{v}_j) \quad \forall i \in \mc{N}$
\end{enumerate}
Note that this is not a true ``distance metric'' since symmetry and the triangle property are not enforced.
The first condition could be extended to include higher order terms.
Let $\vdelta_{ij} = \v{v}_i - \v{v}_j$ for all $i,j \in \mc{N}$.
Finding such a distance metric can be expressed as the following optimization problem:
\begin{subequations} \label{model:1}
\begin{alignat}{4}
\max_{\lambda,\v{a},\v{B},\v{d}}~~& \lambda &&  \\
\st~~& \min_{j \in \mc{C}_i} d_{ij} + \lambda \leq \min_{k \notin \mc{C}_i} d_{ik} & & \qquad \forall i \in \mc{N} \label{eq:separation_requirement} \\
    & d_{ij} = \v{a}^{\top}\vdelta_{ij} + \vdelta_{ij}^{\top}\v{B}\vdelta_{ij} && \qquad \forall i,j \in \mc{N} \\
    & d_{ij} \geq 0 && \qquad  \forall i,j \in \mc{N} \\
    & \v{a} \in \Re^p && \\ 
    & \v{B} \in \Re^{p \times p}, \text{symmetric} &&
\end{alignat}
\end{subequations}

@@

%\section{A linear programming approach to a Feasibility Variant of Model \eqref{model:1}}
\section{A linear programming approach to determine the existence of a feasible distance metric}

The existence of a distance metric satisfying conditions 1-3 can be expressed as the following feasibility (optimization) problem, where $\epsilon > 0$ is a given parameter:
\begin{subequations} \label{model:1_feasibility}
\begin{alignat}{4}
\max_{\lambda,\v{a},\v{B},\v{d}}~~& 0 &&  \\
\st~~& \min_{j \in \mc{C}_i} d_{ij} + \epsilon \leq \min_{k \notin \mc{C}_i} d_{ik} & & \qquad \forall i \in \mc{N} \label{eq:separation_requirement_eps} \\
    & d_{ij} = \v{a}^{\top}\vdelta_{ij} + \vdelta_{ij}^{\top} \v{B}\vdelta_{ij} && \qquad \forall i,j \in \mc{N} \\
    & d_{ij} \geq 0 && \qquad  \forall i,j \in \mc{N} \\
    & \v{a} \in \Re^p && \\ 
    & \v{B} \in \Re^{p \times p}, \text{symmetric} &&
\end{alignat}
\end{subequations}

Given $\epsilon > 0$, model \eqref{model:1} can be expressed as a linear programming feasibility problem, albeit potentially a very large one.
\begin{subequations} \label{model:big_lp_1}
\begin{alignat}{4}
\max_{\v{a},\v{B},\v{d},\v{t}}~~& \sum_{i \in \mc{N}} t_i &&  \\
\st~~& t_i + \epsilon \leq d_{ik} && \qquad \forall i \in \mc{N}, k \in \mc{N}\backslash\mc{C}_i \label{eq:big_lp_separation_constr_t} \\
    & t_i \leq d_{ij} && \qquad \forall i \in \mc{N}, j \in \mc{C}_i \label{eq:big_lp_dual_constr_t} \\
    & d_{ij} = \sum_{k=1}^p \delta_{ijk} a_{k} + \sum_{k=1}^p \delta_{ijk}^2 b_{kk} + 2 \sum_{k=1}^{p-1} \sum_{\ell=k+1}^p  \delta_{ijk} \delta_{ij\ell} b_{k\ell} && \qquad \forall i,j \in \mc{N} \label{eq:big_lp_distance_def_constr_t} \\
    & d_{ij} \geq 0 && \qquad \forall i,j \in \mc{N} \\
    & \v{a} \in \Re^p && \\ 
    & b_{k\ell} \in \Re && \qquad \forall k=1,\dots,p, \ell=k,\dots,p \\
    & t_i \geq 0 & & \qquad \forall i \in \mc{N}
\end{alignat}
\end{subequations}
Model \eqref{model:big_lp_1} is always feasible since the origin $\v{0}$ is always feasible.
Note that $t_i$ represents the distance to point $i$'s nearest neighbor since constraints \eqref{eq:big_lp_dual_constr_t} ensure that $t_i \leq \min_{j \in \mc{C}_i} d_{ij}$,
while the objective function encourages $t_i = \min_{j \in \mc{C}_i} d_{ij}$ for all $i$.
Meanwhile, constraints \eqref{eq:big_lp_separation_constr_t} ensure that the separation $[\min_{k \in \mc{N}\backslash\mc{C}_i} d_{ik}] - [\min_{j \in \mc{C}_i} d_{ij}]$ is at least $\epsilon$.
If model \eqref{model:big_lp_1} returns a feasible solution with $t_i^* = \min_{j \in \mc{C}_i} d_{ij}^* > 0$ for all $i \in \mc{N}$, then the distance metric satisfies requirements 1-3.
Otherwise there exists a subset $\mc{O}$ of outlier points such that $0 \leq t_i^* < \min_{j \in \mc{C}_i} d_{ij}^*$ for all $i \in \mc{O}$, which means that no such distance metric guaranteeing a minimum separation exists for the choice of $\epsilon$.

\section{Mixed-Integer Linear Optimization Approaches}

@@

>>>>

\newcommand{\xor}{\underline{\vee}}
\newcommand{\Xor}{\underline{\bigvee}}

@@

Suppose we are given $N$ points $\v{v}_i \in \Re^p$ with class membership $\mc{C}_i$ for $i \in \mc{N} = \{1,\dots,N\}$.
Our goal is to find a distance metric $\D: \Re^p \mapsto \Re$ such that the following conditions hold:
\begin{enumerate}
\item $\D(\v{v}_i,\v{v}_j) = \v{a}^{\top}(\v{v}_i - \v{v}_j) + (\v{v}_i - \v{v}_j)^{\top} \v{B}(\v{v}_i - \v{v}_j) \quad \forall i,j \in \mc{N}$, for some $\v{a} \in \Re^p$ and $\v{B} \in \Re^{p \times p}$ and symmetric
\item $\D(\v{v}_i,\v{v}_j) \geq 0 \quad \forall i,j \in \mc{N}$
\item $\min_{j \in \mc{C}_i} \D(\v{v}_i,\v{v}_j) < \min_{k \notin \mc{C}_i} \D(\v{v}_i,\v{v}_j) \quad \forall i \in \mc{N}$
\end{enumerate}
Note that this is not a true ``distance metric'' since symmetry and the triangle property are not enforced.
The first condition could be extended to include higher order terms.
Let $\vdelta_{ij} = \v{v}_i - \v{v}_j$ for all $i,j \in \mc{N}$.
Finding such a distance metric can be expressed as the following optimization problem:

\begin{subequations} \label{model:1}
\begin{alignat}{4}
\max_{\lambda,\v{a},\v{B},\v{d}}~~& \lambda &&  \\
\st~~& \min_{j \in \mc{C}_i} d_{ij} + \lambda \leq \min_{k \notin \mc{C}_i} d_{ik} & & \qquad \forall i \in \mc{N} \label{eq:separation_requirement} \\
    & d_{ij} = \v{a}^{\top}\vdelta_{ij} + \vdelta_{ij}^{\top}\v{B}\vdelta_{ij} && \qquad \forall i,j \in \mc{N} \\
    & d_{ij} \geq 0 && \qquad  \forall i,j \in \mc{N} \\
    & \v{a} \in \Re^p && \\ 
    & \v{B} \in \Re^{p \times p}, \text{symmetric} &&
\end{alignat}
\end{subequations}

@@

\newpage
%\section{A linear programming approach to a Feasibility Variant of Model \eqref{model:1}}
\section{A linear programming approach to determine the existence of a feasible distance metric}

Let $\mc{F}$ denote the feasible region associated with the distance metric parameters, i.e.,
\begin{equation}
\mc{F} = \{ (\v{a},\v{B},\v{d}) \in \Re^p \times \Re^{p \times p} \times \Re_+^{N \times N} : \v{B}~\text{symmetric}, d_{ij} = \v{a}^{\top}\vdelta_{ij} + \vdelta_{ij}^{\top} \v{B}\vdelta_{ij} \quad \forall i,j \in \mc{N} \}
\end{equation}
or
\begin{subequations} \label{set:feasible_region_distance_metric}
\begin{alignat}{4}
\mc{F} = \Bigg\{ \v{a},\v{B},\v{d} : &  &&  \\
    & d_{ij} = \sum_{k=1}^p \delta_{ijk} a_{k} + \sum_{k=1}^p \delta_{ijk}^2 b_{kk} + 2 \sum_{k=1}^{p-1} \sum_{\ell=k+1}^p  \delta_{ijk} \delta_{ij\ell} b_{k\ell} && \qquad \forall i,j \in \mc{N} \label{eq:distance_metric_def} \\
    & 0 \leq d_{ij} \leq 1 && \qquad \forall i,j \in \mc{N} \\
    & \v{a} \in \Re^p && \\ 
    & b_{k\ell} \in \Re && \qquad \forall k=1,\dots,p, \ell=k,\dots,p \Bigg\}
\end{alignat}
\end{subequations}

The existence of a distance metric satisfying conditions 1-3 can be expressed as the following feasibility (nonlinear optimization) problem, where $\epsilon > 0$ is a given parameter:
%\begin{subequations} \label{model:1_feasibility}
%\begin{alignat}{4}
%\max_{\v{a},\v{B},\v{d}}~~& 0 &&  \\
%\st~~& \min_{j \in \mc{C}_i} d_{ij} + \epsilon \leq \min_{k \notin \mc{C}_i} d_{ik} & & \qquad \forall i \in \mc{N} \label{eq:separation_requirement_eps} \\
%    & d_{ij} = \v{a}^{\top}\vdelta_{ij} + \vdelta_{ij}^{\top} \v{B}\vdelta_{ij} && \qquad \forall i,j \in \mc{N} \\
%    & d_{ij} \geq 0 && \qquad  \forall i,j \in \mc{N} \\
%    & \v{a} \in \Re^p && \\ 
%    & \v{B} \in \Re^{p \times p}, \text{symmetric} &&
%\end{alignat}
%\end{subequations}
\begin{subequations} \label{model:1_feasibility}
\begin{alignat}{4}
\max_{\v{a},\v{B},\v{d}}~~& 0 &&  \\
\st~~& \min_{j \in \mc{C}_i} d_{ij} + \epsilon \leq \min_{k \notin \mc{C}_i} d_{ik} & & \qquad \forall i \in \mc{N} \label{eq:separation_requirement_eps} \\
    & (\v{a},\v{B},\v{d}) \in \mc{F} &&
\end{alignat}
\end{subequations}
Note that if we constrain the distances to be bounded (e.g., $d_{ij} \in [0,1]$), problem \eqref{model:1_feasibility} is not guaranteed to be feasible for an arbitrary choice of $\epsilon$. 
To see this, let $\epsilon > 0$ and consider an example with three points and two classes.  Point 1 is located at (0,0), point 2 at (1,1), and point 3 at $(\epsilon/2,\epsilon/2)$.  Points 1 and 2 belong to class 1, while point 3 belongs to class 2. There is no way to create a distance metric that guarantees that $d_{13} > \epsilon$ and $d_{23} > \epsilon$.

There are several ways to proceed.

\subsection{Multi-Objective approach}
Let $W$ be a positive scalar.  
We could solve the following LP that maximizes both $\epsilon$ and the minimum distance variables $t_i$:
\begin{subequations} \label{model:multi_objective_lp_1}
\begin{alignat}{4}
\max_{\v{a},\v{B},\v{d},\epsilon,\v{t}}~~& W\epsilon + \sum_{i \in \mc{N}} t_i &&  \\
\st~~& t_i + \epsilon \leq d_{ik} && \qquad \forall i \in \mc{N}, k \in \mc{N}\backslash\mc{C}_i \label{eq:mo_lp_separation_constr_t} \\
    & t_i \leq d_{ij} && \qquad \forall i \in \mc{N}, j \in \mc{C}_i \label{eq:mo_lp_dual_constr_t} \\
    & t_i \geq 0 & & \qquad \forall i \in \mc{N} \\
    & \eqref{set:feasible_region_distance_metric} & & 
\end{alignat}
\end{subequations}
\begin{proposition}
Problem \eqref{model:multi_objective_lp_1} is always feasible.
\end{proposition}
\proof The origin is always a feasible solution.  That is, set $\v{a} = \v{0},\v{B} = \v{0},\v{d} = \v{0},\epsilon = 0,\v{t} = \v{0}$.
\qed

\subsection{Iterative LP approach}

First solve the following problem to determine the maximum non-neighbor separation $\epsilon$ that is possible. 
\begin{subequations} \label{model:epsilon_lp_1}
\begin{alignat}{4}
\max_{\v{a},\v{B},\v{d},\epsilon}~~& \epsilon &&  \\
\st~~& \epsilon \leq d_{ik} && \qquad \forall i \in \mc{N}, k \in \mc{N}\backslash\mc{C}_i \label{eq:epsilon_lp_separation_constr_t} \\
    & \eqref{set:feasible_region_distance_metric} & & 
\end{alignat}
\end{subequations}
Let $\epsilon^{\max}$ be the maximum non-neighbor separation, i.e., the optimal value of $\epsilon$ in \eqref{model:epsilon_lp_1}.
 Note that this problem is identical to problem \eqref{model:multi_objective_lp_1} after omitting the $\v{t}$ decision variables.  Hence, it is guaranteed to be feasible.

Given $\epsilon \in (0,\epsilon^{\max}]$, model \eqref{model:1} can be expressed as a linear programming feasibility problem:
\begin{subequations} \label{model:big_lp_1}
\begin{alignat}{4}
\max_{\v{a},\v{B},\v{d},\v{t}}~~& \sum_{i \in \mc{N}} t_i &&  \\
\st~~& t_i + \epsilon \leq d_{ik} && \qquad \forall i \in \mc{N}, k \in \mc{N}\backslash\mc{C}_i \label{eq:big_lp_separation_constr_t} \\
    & t_i \leq d_{ij} && \qquad \forall i \in \mc{N}, j \in \mc{C}_i \label{eq:big_lp_dual_constr_t} \\
    & t_i \geq 0 & & \qquad \forall i \in \mc{N} \\
    & \eqref{set:feasible_region_distance_metric} & & 
\end{alignat}
\end{subequations}
For $\epsilon \in (0,\epsilon^{\max}]$, model \eqref{model:big_lp_1} is always feasible since the origin $\v{0}$ is always feasible.
Note that $t_i$ represents the distance to point $i$'s nearest neighbor since constraints \eqref{eq:big_lp_dual_constr_t} ensure that $t_i \leq \min_{j \in \mc{C}_i} d_{ij}$,
while the objective function encourages $t_i = \min_{j \in \mc{C}_i} d_{ij}$ for all $i$.
Meanwhile, constraints \eqref{eq:big_lp_separation_constr_t} ensure that the separation $[\min_{k \in \mc{N}\backslash\mc{C}_i} d_{ik}] - [\min_{j \in \mc{C}_i} d_{ij}]$ is at least $\epsilon$.
If model \eqref{model:big_lp_1} returns a feasible solution with $t_i^* = \min_{j \in \mc{C}_i} d_{ij}^* > 0$ for all $i \in \mc{N}$, then the distance metric satisfies requirements 1-3.
Otherwise there exists a subset $\mc{O}$ of outlier points such that $0 \leq t_i^* < \min_{j \in \mc{C}_i} d_{ij}^*$ for all $i \in \mc{O}$, which means that no such distance metric guaranteeing a minimum separation exists for the choice of $\epsilon$.

The following algorithm can be used to explore potential distance metrics for different levels of separation.

\begin{algorithm}
%{\small
\caption{Iterative LP Heuristic}
\label{algo:iterative_LP_heuristic}
\begin{algorithmic}[1]
\STATE Solve model \eqref{model:epsilon_lp_1}. Let $\epsilon^{\max}$ be the optimal objective function value.
\STATE Let $\mc{E} = \{ i / \epsilon^{\max} : i=1,\ldots,I \}$ (e.g., $I=10$)
\FOR{ $\epsilon \in \mc{E}$ }
	\STATE Solve model \eqref{model:big_lp_1}.
	\STATE Store the outliers and distance metric
\ENDFOR
\end{algorithmic}
%}
\end{algorithm}

\newpage
\section{Mixed-Integer Linear Optimization Approaches}

@@

\newpage
\section{Generalized disjunctive programming (GDP) formulations}

This section presents several generalized disjunctive programming formulations.

\textbf{Robust GDP formulation with no outliers}
\begin{subequations} \label{model:gdp_robust_no_outliers}
\begin{alignat}{4}
\max_{\lambda,\v{a},\v{B},\v{d}}~~& \lambda &&  \\
\st~~& \mathop{\Xor}_{j \in \mc{C}_i} [ d_{ij} + \lambda \leq d_{ik} \quad \forall k \notin \mc{C}_i ] && \qquad \forall i \in \mc{N} \label{eq:gdp_separation_constr} \\
    & (\v{a},\v{B},\v{d}) \in \mc{F} && 
\end{alignat}
\end{subequations}
Constraints \eqref{eq:gdp_separation_constr} read: ``the distance between point $i$ and its in-class neighbor $j_1 \in \mc{C}_i$ plus the separation $\lambda$ must not exceed the distance between point $i$ and all of its non-neighbors'' or ``the distance between point $i$ and its in-class neighbor $j_2 \in \mc{C}_i$ plus $\lambda$ must not exceed ... '' or ...

\textbf{Robust GDP formulation with outliers}

\begin{subequations} \label{model:gdp_robust_with_outliers}
\begin{alignat}{4}
\max_{\lambda,\v{a},\v{B},\v{d}}~~& \lambda - \rho \sum_{i \in \mc{N}} z_i &&  \\
\st~~& \Bigg[ \mathop{\Xor}_{j \in \mc{C}_i} [ d_{ij} + \lambda \leq d_{ik} \quad \forall k \notin \mc{C}_i ] \Bigg]~\mathop{\Xor}~[z_i = 1] && \qquad \forall i \in \mc{N} \label{eq:gdp_robust_with_outliers_separation_constr} \\
    & (\v{a},\v{B},\v{d}) \in \mc{F}, \v{z} \in \Re_+^N && 
\end{alignat}
\end{subequations}

\textbf{GDP formulation with outliers}

\begin{subequations} \label{model:gdp_nonrobust_with_outliers}
\begin{alignat}{4}
\max_{\vlambda,\v{a},\v{B},\v{d}}~~& \sum_{i \in \mc{N}} \lambda_i &&  \\
\st~~& \Bigg[ \mathop{\Xor}_{j \in \mc{C}_i} [ d_{ij} + \lambda_i \leq d_{ik} \quad \forall k \notin \mc{C}_i; \lambda_i \geq 0 ] \Bigg]~\mathop{\Xor}~[\lambda_i = -\rho] && \qquad \forall i \in \mc{N} \label{eq:gdp_nonrobust_with_outliers_separation_constr} \\
    & (\v{a},\v{B},\v{d}) \in \mc{F}, \lambda_i \geq -\rho && 
\end{alignat}
\end{subequations}
